---
title: "Final Assignment R"
author: "Manoj Srivathsava"
format:
  pdf:
    code-overflow: wrap         # Automatically wrap long code lines
    code-block-font-size: 0.9em
    geometry: "margin=1in"     # Set 1-inch margins
    include-in-header:
      text: |
        % Allow line breaks in the code/highlighting environment
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
editor: visual
date: "`r Sys.Date()`"
execute:
  echo: true
  message: false
  warning: false
---

# Introduction

This project analyses vehicle licensing data in Ireland using techniques covered in the *Data Programming with R* module.\
The dataset contains a mix of categorical and numerical variables, making it suitable for data cleaning, exploratory analysis, functional programming, and custom function development.

------------------------------------------------------------------------

# Part 1: Data Analysis

## 1. Libraries

```{r message=FALSE, warning=FALSE}
# Core packages for data manipulation and visualisation
library(tidyverse)
library(ggplot2)

# Packages for data cleaning and summaries
library(janitor)
library(skimr)
library(knitr)

# Functional programming
library(purrr)
```

These libraries support efficient data handling, clear visualisation, and reproducible analysis.

------------------------------------------------------------------------

## 2. Data Loading

```{r}
# Load the dataset
data <- read.csv("DOTA04.20251216T121242.csv")

# Check dataset dimensions
dim(data)
```

The dataset contains **3,240 observations and 8 variables**, providing sufficient scope for meaningful exploratory analysis.

------------------------------------------------------------------------

## 3. Data Preprocessing

```{r}
# Convert relevant variables to appropriate types and remove missing values
data_clean <- data |>
  mutate(
    Statistic.Label = as.factor(Statistic.Label),
    Fuel.Type = as.factor(Fuel.Type),
    Type.of.Carriage = as.factor(Type.of.Carriage),
    Licensing.Authority = as.factor(Licensing.Authority),
    Gross.Vehicle.Weight = as.factor(Gross.Vehicle.Weight),
    Year = as.integer(Year)
  ) |>
  filter(!is.na(VALUE))
```

Categorical variables are converted to factors, and missing values in the response variable are removed to ensure reliable summaries and plots.

------------------------------------------------------------------------

## 4. Numerical Summaries

```{r}
# Summary statistics by fuel type
fuel_summary <- data_clean |>
  group_by(Fuel.Type) |>
  summarise(
    mean_value = mean(VALUE),
    median_value = median(VALUE),
    total_value = sum(VALUE),
    .groups = "drop"
  )

# Summary by licensing authority
authority_summary <- data_clean |>
  group_by(Licensing.Authority) |>
  summarise(
    total_value = sum(VALUE),
    .groups = "drop"
  )

fuel_summary
authority_summary
```

These summaries highlight differences in vehicle counts across fuel types and regions.

------------------------------------------------------------------------

## 5. Visual Exploration

### 5.1 Vehicles by Licensing Authority

```{r}
ggplot(authority_summary,
       aes(x = reorder(Licensing.Authority, total_value), y = total_value)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Total Vehicles by Licensing Authority",
    x = "Licensing Authority",
    y = "Total Vehicles"
  ) +
  theme_minimal()
```

This plot reveals substantial regional variation in vehicle registrations.

------------------------------------------------------------------------

### 5.2 Distribution by Fuel Type

```{r}
ggplot(data_clean, aes(x = Fuel.Type, y = VALUE)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(
    title = "Distribution of Vehicle Counts by Fuel Type",
    x = "Fuel Type",
    y = "Vehicle Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The boxplot shows differences in spread and the presence of outliers across fuel types.

------------------------------------------------------------------------

### 5.3 Trend Over Time (Diesel Vehicles)

```{r}
ggplot(
  data_clean |> filter(Fuel.Type == "Diesel"),
  aes(x = Year, y = VALUE)
) +
  stat_summary(fun = sum, geom = "line", linewidth = 1, color = "steelblue") +
  labs(
    title = "Trend in Diesel Vehicle Counts Over Time",
    x = "Year",
    y = "Total Vehicles"
  ) +
  theme_minimal()
```

This line plot highlights changes in diesel vehicle registrations over time.

------------------------------------------------------------------------

### 5.4 Fuel Composition by Region

```{r}
ggplot(data_clean, aes(x = Licensing.Authority, y = VALUE, fill = Fuel.Type)) +
  stat_summary(fun = sum, geom = "bar", position = "stack") +
  labs(
    title = "Fuel Type Composition by Licensing Authority",
    x = "Licensing Authority",
    y = "Total Vehicles"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

This stacked bar chart shows how fuel types contribute to total registrations within each region.

------------------------------------------------------------------------

## 6. Functional Programming with purrr

```{r}
# Split data by licensing authority
authority_list <- split(data_clean, data_clean$Licensing.Authority)

# Apply the same summary function to each subset
authority_purrr_summary <- map_dfr(
  authority_list,
  ~ summarise(.x,
              mean_value = mean(VALUE),
              total_value = sum(VALUE)),
  .id = "Licensing.Authority"
)

authority_purrr_summary
```

Using `purrr::map_dfr()` allows the same operation to be applied consistently across groups.

------------------------------------------------------------------------

# Part 2: R Package Demonstration

```{r}
# Clean column names
clean_data <- data |>
  clean_names()

# Frequency table by fuel type
fuel_table <- clean_data |>
  tabyl(fuel_type)

fuel_table |>
  adorn_totals("row") |>
  adorn_percentages("col") |>
  adorn_pct_formatting()

# Package citation
citation("janitor")
```

The `janitor` package simplifies data cleaning and produces clear, publication-ready tables.

------------------------------------------------------------------------

# Part 3: Functions and Programming

## 1. Custom Analysis Function

```{r}
vehicle_group_analysis <- function(data, group_var, value_var) {

  # Remove missing values
  clean_data <- data[!is.na(data[[value_var]]), ]

  # Compute group-wise statistics
  stats <- tapply(
    clean_data[[value_var]],
    clean_data[[group_var]],
    function(x) {
      c(
        mean = mean(x),
        median = median(x),
        sd = sd(x),
        n = length(x)
      )
    }
  )

  # Create S3 object
  output <- list(
    statistics = as.data.frame(do.call(rbind, stats)),
    group_var = group_var,
    value_var = value_var
  )

  class(output) <- "vehicle_analysis"
  output
}
```

------------------------------------------------------------------------

## 2. S3 Methods

```{r}
print.vehicle_analysis <- function(x, ...) {
  cat("Vehicle Group Analysis Object\n")
  cat("Grouping variable:", x$group_var, "\n")
  cat("Value variable:", x$value_var, "\n")
  cat("Number of groups:", nrow(x$statistics), "\n")
}

summary.vehicle_analysis <- function(object, ...) {
  cat("Detailed Group-wise Statistics\n\n")
  print(round(object$statistics, 2))
}

plot.vehicle_analysis <- function(x, ...) {
  barplot(
    x$statistics$mean,
    names.arg = rownames(x$statistics),
    col = "steelblue",
    las = 2,
    main = paste("Mean", x$value_var, "by", x$group_var),
    ylab = paste("Mean", x$value_var)
  )
}
```

------------------------------------------------------------------------

## 3. Example Usage

```{r}
vehicle_result <- vehicle_group_analysis(
  data = data,
  group_var = "Fuel.Type",
  value_var = "VALUE"
)

print(vehicle_result)
summary(vehicle_result)
plot(vehicle_result)
```

------------------------------------------------------------------------

# Conclusion

This project demonstrates the application of data cleaning, exploratory analysis, functional programming, and custom S3 classes in R.\
The results provide insight into vehicle registration patterns across fuel types and regions while showcasing reusable and well-documented code.
